Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-01-31 paper/paper.tex
made cahnges to first parts of paper.  Havent yet done methods-pill section because want to know narrative first Tested everything on node08.

\section{Introduction}
It is vital that during the process of bacterial cell division, a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  A long FtsZ polymer chain develops on the cell wall, in the
center region of the cell, the region which dictates the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer\cite{shen2010examination}, exhibits pole to
pole oscillatory behavior in conjunction with the MinD and MinE
proteins, while the oscillating MinE tends to be more localized in the
center of the cell~\cite{hu1999topological, fu2001mine,
  shapiro2009and, yu1999ftsz, raskin1999rapid}. The MinC protein will
then have a higher time averaged concentration in the cell poles, as
opposed to the center region of the cell, aiding in prohibiting the
FtZ from developing in the wrong region.

The cell division process must be able to prevent minicelling, even
under irregular conditions. Studies have shown that it is capable of
performing this task for fairly perturbed cell
shapes\cite{touhami2006temperature}. Mannik et al. have done some
early experimental work showing regular oscillations of cell division
systems in bacteria that were physically forced into unusual
shapes. Varma et al. studied one particular example of a perturbed
shape, a three-pronged tube, and found both experimental and
simulation results\cite{varma2008min} supporting the idea that certain
systems inside the cell begin to undergo oscillations inside the
cell. These oscillations seem to seek out the extreme poles in the
cell shape\cite{corbin2002exploring} \cite{juarez2010changes}, and
effectively prevent minicelling in most cases. However, Mannik et
al. have also shown that there are limitations to this
mechanism\cite{mannik2010bacteria} \cite{mannik2009bacterial}. One
instance of this behavior is observable when the cell has been forced
into a flattened, irregular shape. Oscillations in this sort of cell
shape are far from the regular sorts of oscillations seen in the
computational work done by Huang et al\cite{huang2003dynamic}. We seek
to apply a computational model to explore the more extreme or
irregular cell shapes, in order to determine the limits of the steady
cell oscillations seen by Mannik et al.

%is this necessary? was in old paragraph.
%\fixme{this link doesn't work =
%  cite{mannik2012robustness}} \cite{mannik2010bacteria}
%\cite{mannik2009bacterial}

A significant amount of work has been done to develop protein reaction
and diffusion models that exhibit accurate macroscopic dynamics of the
MinD protein system. Early models involved free proteins that affect
each others' rates of diffusion and membrane attachment, but do not
combine into compound states~\cite{meinhardt2001pattern}.  In 2003
Huang improved upon this work with a simple model based on MinD-MinE
combination, ATPase hydrolysis, and MinD membrane attachment that when
simulated, exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is more
likely to attach to the membrane when MinD is already clustered there
(following observed non-linear attachement of minD on the cell
membrane), and is stationary once attached.  A number of studies have
used an approach similar in that they do not rely on the ability of
MinD to move along the walls and
cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.  Variations of the Huang 2003 model that
stochasticly account for variations of molecular interaction
\cite{fange2006noise} and monte-carlo simulations that implement this
stochastic version of Huang's mean field reaction rates confirm the
major results obtained by Huang's model, and more successfully predict
experimentally observed oscillations in round cell
phenotypes~\cite{drew2005polymerization, fange2006noise,
  huang2004min}. Biochemical models of broader scope have also been
used to study the MinD system and show consistent
results.\cite{arjunan2010new}.  However, in general, the results of the
stochastic and monte-carlo simulations are similar to those given by
Huang's mean field results.

@@

\section{Model and Methods}
We implement the model of Huang \emph{et al.}  for the behavior of the
MinD and MinE proteins inside the cell~\cite{huang2003dynamic}.
Figure~\ref{fig:reactions} shows the reaction process.  The
cytoplasmic MinD:ADP complex undergoes nucleotide exchange and is
changed into the MinD:ATP complex.  This will naturally diffuse
towards and attach to the cell membrane.  A cytoplasmic MinE will
attach to the wall bound MinD:ATP complex and after a time will
activate ATP hydrolosis.  This breaks up the complex, releasing MinE,
phosphate, and MinD:ADP back into the cell.  The MinD:ADP will undergo
nucleotide exchange and begin again the cyclic process.  This model is
defined by a set of five reaction-diffusion equations:

@@

Our diffusion and reaction rates are shown below.  We are interested
primarily in the effect of cellular shape on the protein
oscillations, so we follow Huang's work\cite{huang2003dynamic} and do
not deviate from the wild type values used in that work.  Huang's
simulations use total MinD and MinE concentrations of $1,000/\mu m$
and $350/\mu m$, respectively, in a cylindrical cell of radius $0.5\mu
m$, and in our (non-cylindrical) cells we use the same number of
proteins per unit volume.  These concentration values are 1273 $\mu
m^{-3}$ and 445 $\mu m^{-3}$, respectively.
\begin{gather*}
  \mathcal{D}_D = \mathcal{D}_{E} = 2.5\micron^2/\text{sec}\\
  k_D^{\textrm{ADP $\rightarrow$ ATP}} = 1/\textrm{sec,  }
  k_D = 0.025 \micron /\textrm{sec}\\
  k_{dD} = 0.0015 \micron^3/ \textrm{sec,  }
  k_{de} = 0.7/\textrm{sec}\\
  k_E = 0.093 \micron^3 /\textrm{sec}.
\end{gather*}
A 3D grid is then constructed in cartesian coordinates, with a grid
spacing of .05\micron. We define our cell shapes and solve the
reaction-diffusion partial differential equations numerically to
observe the time evolution of the MinD and MinE concentrations inside
the cell.

@@

\section{Cell shapes}

@@

Kubitschek has shown in multiple experiments that at the time of cell
division cells have a volume that is within a range of roughly
$1\micron^3$ to $2\micron^3$~\cite{kubitschek1990cell,
  kubitschek1968linear}.  We follow Huang's
simulations\cite{huang2003dynamic} and Mannik's experiments and model
cells that are slightly larger than this range.  Amoung our flattened
cells the two dimensional length scales are tuned so that total cell
volumes are very close to $3\micron^3$. \jeffsays{I'm not sure how
  true this is.  We should go through and check.  There've been a few
  implementations of printouts reporting volume in the cpp}
\section{Specific Results}

\subsection{Pill Shape}

\begin{figure*}
  \includegraphics[width=\textwidth]{../data/shape-p/plots/image-plot--p-300-50-0-0-1500.pdf}
  \caption{Images of the concentration of each species in a natural
    pill-shaped bacterium as a function of time, with darker regions
    indicating higher concentration.  The cells pictured are
    $4\micron$ in length, measured from end to end.  Frames from an animation showing
    a series of contour plots of the concentration of proteins in
    different stages over time.  The order of frames is such that
    individual MinD proteins begin at the bottom of the plot (in the
    MinD:ATP state in the cytoplasm), and progress upward until they
    reach the MinE:MinD:ATP membrane-bound complex.  At that point,
    they will spontaneously dissociate into cytoplasmic MinE (the top
    row) and the starting state of cytoplasmic MinD:ADP.  The
    densities plotted are integrated along the axis orthogonal to the
    page, and the color scale is chosen separately for each species
    with black as the maximum value over space and time.}
  \label{image-p}
\end{figure*}

@@

@@

\subsubsection{Randst-mannik shape}
\begin{figure}
  \includegraphics[width=2.8cm]{../mannik-2.png}
  \includegraphics[width=2.8cm]{../mannik-1.png}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-1300-2000-9400-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-1101-1101-9500-1500.pdf}
  \caption{Arrow plots for our simulation and mannik's experimental}
  \label{arrow-compare-mannik}
\end{figure}

Figure \ref{arrow-compare-mannik} shows a comparison of time maxima
plots with similar plots published by mannik et al.  The sequence of
arrows show a series of global maxima in space and local maxima in
time at their heads, numbered by the sequence in which they occur.
They function as a trace of the path that the traveling maxima take.
We have created for each of the two shapes published by Mannik a
closely resembling shape (shown directly below Mannik's) and a
variation on this shape (shown below that) in order to test the effect
of variations on basic shapes.  The plots of the left shape show that
much of the preffered locations for the proteins to maximize are
similar between this experimental and simulated shape.  The maxima in
the top center of the cell does not seem to appear in the experimental
version.  Also in both of the simulations on the left there don't seem
to be maxima at membrane regions with negative radii of curvature
while the experimental shows a maxima in the bottom center region
right at one of these regions.  The shape on the right compares less
successfully.  Here the simulation shows a limited number of maxima
locations that appear to be quite regular, where mannik's experimental
results show a much more chaotic system.  In particular, the maxima
location on the right wall seems to be very robust.  It acts as one of
the poles do in the pill shapes.  \fixme{I'm now running a sim with a
  slightly different shape to see result} The simulated version has
found a robust two pole pattern, similar to that seen in other shapes.

>>>>

\section{Introduction}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer\cite{shen2010examination}, exhibits pole to
pole oscillatory behavior in conjunction with the MinD protein and a
MinE that tends to be more localized in the center of the
cell~\cite{hu1999topological, fu2001mine, shapiro2009and, yu1999ftsz,
  raskin1999rapid}. The MinC protein will then have a higher time
averaged concentration in the cell poles, as opposed to the center
region of the cell, aiding in prohibiting the FtZ from developing in
the wrong region.

Studies have shown that the cell division process is capable of
preventing minicelling for fairly perturbed cell
shapes\cite{touhami2006temperature}. Varma et al. studied one
particular example of a perturbed shape, a three-pronged tube, and
found both experimental and simulation results\cite{varma2008min} that
show oscillation. These oscillations seem to seek out the extreme
poles in the cell shape\cite{corbin2002exploring}
\cite{juarez2010changes}, and effectively prevent minicelling in most
cases. However, Mannik et al. have also shown that there are
limitations to this mechanism\cite{mannik2010bacteria}
\cite{mannik2009bacterial}. One instance of this behavior is
observable when the cell has been forced into a flattened, irregular
shape. Oscillations in this sort of cell shape are far from the
regular sorts of oscillations seen in the computational work done by
Huang et al\cite{huang2003dynamic}. We seek to apply a computational
model to explore the more extreme or irregular cell shapes, in order
to determine the limits of the steady cell oscillations seen by both
theory and experiment.

A significant amount of work has been done to develop protein reaction
and diffusion models that exhibit accurate macroscopic dynamics of the
MinD protein system. Early models involve free proteins that affect
each others' rates of diffusion and membrane attachment, but do not
combine into compound states~\cite{meinhardt2001pattern}.  In 2003
Huang improved upon this work with a simple simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in
cylindrical cells~\cite{huang2003dynamic}. In this model cytoplasmic
MinD is more likely to attach to the membrane when MinD is already
clustered there (following observed non-linear attachement of minD on
the cell membrane), and is stationary once attached.  A number of
studies have used an approach similar in that they do not rely on the
ability of MinD to move along the walls and
cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.  Variations of the Huang 2003 model that
stochasticly account for variations of molecular interaction
\cite{fange2006noise} and monte-carlo simulations that implement this
stochastic version of Huang's mean field reaction rates confirm the
major results obtained by Huang's model, and more successfully predict
experimentally observed oscillations in round cell
phenotypes~\cite{drew2005polymerization, fange2006noise,
  huang2004min}. Biochemical models of broader scope have also been
used to study the MinD system and show consistent
results.\cite{arjunan2010new}.  However, in general, the results of
the stochastic and monte-carlo simulations are similar to those given
by Huang's mean field results.

@@

\section{Model, Methods, and Pill Shaped Cell}
We implement the model of Huang \emph{et al.}  for the behavior of the
MinD and MinE proteins inside the cell~\cite{huang2003dynamic}.
Figure~\ref{fig:reactions} shows the reaction process.  The
cytoplasmic MinD:ADP complex undergoes nucleotide exchange and is
changed into the MinD:ATP complex.  This will naturally diffuse
towards and attach to the cell membrane.  A cytoplasmic MinE will
attach to the wall bound MinD:ATP complex and after a time will
activate ATP hydrolosis.  This breaks up the complex, releasing MinE,
phosphate, and MinD:ADP back into the cytoplasm.  The MinD:ADP will undergo
nucleotide exchange and begin again the cyclic process.  This model is
defined by a set of five reaction-diffusion equations:

@@

Our diffusion and reaction rates are shown below.  We are interested
primarily in the effect of cellular size and shape on the protein
oscillations, so we follow Huang\cite{huang2003dynamic} and do not
deviate from the wild type values used in the cited work (see values
below).  Huang's simulations use total MinD and MinE concentrations of
$1,000/\mu m$ and $350/\mu m$, respectively, in a cylindrical cell of
radius $0.5\mu m$, and in our (non-cylindrical) cells we use the same
number of proteins per unit volume.  These concentration values are
1273 $\mu m^{-3}$ and 445 $\mu m^{-3}$, respectively.
\begin{gather*}
  \mathcal{D}_D = \mathcal{D}_{E} = 2.5\micron^2/\text{sec}\\
  k_D^{\textrm{ADP $\rightarrow$ ATP}} = 1/\textrm{sec,  }
  k_D = 0.025 \micron /\textrm{sec}\\
  k_{dD} = 0.0015 \micron^3/ \textrm{sec,  }
  k_{de} = 0.7/\textrm{sec}\\
  k_E = 0.093 \micron^3 /\textrm{sec}.
\end{gather*}
A 3D grid is then constructed in cartesian coordinates, with a grid
spacing of .05\micron. We define our cell shapes and solve the
reaction-diffusion partial differential equations numerically to
observe the time evolution of the MinD and MinE concentrations inside
the cell.

@@

@@

Kubitschek has shown in multiple experiments that at the time of cell
division cells have a volume that is within a range of roughly
$1\micron^3$ to $2\micron^3$~\cite{kubitschek1990cell,
  kubitschek1968linear}.  We follow Huang's
simulations\cite{huang2003dynamic} and Mannik's experiments and model
cells that are slightly larger than this range.  Amoung our flattened
cells the two dimensional length scales are tuned so that total cell
volumes are very close to $3\micron^3$. \jeffsays{I'm not sure how
  true this is.  We should go through and check.  There've been a few
  implementations of printouts reporting volume in the cpp}

\begin{figure*}
  \includegraphics[width=\textwidth]{../data/shape-p/plots/image-plot--p-300-50-0-0-1500.pdf}
  \caption{Countour plot images of the concentration of each protein
    species in a natural pill-shaped bacterium at regular intervals in
    time (one second intervals), with darker regions indicating higher
    concentration.  The cells pictured are $4\micron$ in length,
    measured from end to end.  The order of frames is such that
    individual MinD proteins begin at the bottom of the plot (in the
    MinD:ATP state in the cytoplasm), and progress upward until they
    reach the MinE:MinD:ATP membrane-bound complex.  At that point,
    they will spontaneously dissociate into cytoplasmic MinE (the top
    row) and the starting state of cytoplasmic MinD:ADP.  The
    densities plotted are integrated along the axis orthogonal to the
    page, and the color scale is chosen separately for each species
    with black as the maximum value over space and time.}
  \label{image-p}
\end{figure*}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{stability-analysis}
  \caption{Infinite slab stability analysis yields characteristic half
    wavelengths at which system becomes unstable.}
    \label{fig:stability-analysis}
\end{figure}

\section{Theory}
Stability analysis of the five differential equations in an infinite
slab show a characteristic half wavelength in perturbative spatial
oscillations at which the system becomes unstable.  This length is
dependent upon slab thickness as shown in
Figure~\ref{fig:stability-analysis}.  When increasing the lengths and
widths of our simulated flattened cells, the steady, two pole
oscillations seem to break down when the the shortest distance across
the cell is approximately equal to the characteristic half wavelength
at the particular thickness at which we simulate (our simulations of cells
with thickness $0.25\mu m$ corresponds to half wavelengths of $1.91\mu
m$).  Huang~\cite{huang2003dynamic} performs a similar analysis for a
cylindrical shape and finds characteristic half wavelengths of $2\mu
m$.

@@

\subsubsection{Randst-mannik shape}
\begin{figure}
  \includegraphics[width=2.8cm]{../mannik-2.png}
  \includegraphics[width=2.8cm]{../mannik-1.png}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-1300-2000-9400-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-1100-1100-9500-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-1800-2500-9400-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-1800-1800-9500-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-2300-3300-9400-1500.pdf}
  \includegraphics[width=3cm]{../data/shape-randst/plots/arrow-plot-NflD-randst-25-2800-2800-9500-1500.pdf}
  \caption{Arrow plots for our simulation and mannik's experimental}
  \label{arrow-compare-mannik}
\end{figure}

Figure~\ref{arrow-compare-mannik} shows a comparison of time maxima
plots with similar plots published by mannik et al.  The sequence of
arrows show a series of global maxima in space and local maxima in
time at their heads, numbered by the sequence in which they occur.
They function as a trace of the path that the traveling maxima take.
We have created for each of the two shapes published by Mannik a
closely resembling shape (shown directly below Mannik's) that is in
all dimensions the same size as that reported in their experiment.  We
have also simulated longer and wider versions of these same shapes
(keeping the cell thickness the same).  Shown in
Figure~\ref{arrow-compare-mannik} are versions that have 1.3 and 2.0
times the length and width dimensions.  These plots show
that there are certain sizes at which the steady, two pole
oscillations seem to break down, and indeed these lengths seem to be
roughly those estimated by our 2D stability analysis.  \fixme{Waiting
  to rewrite this section when more data, at more sizes, becomes
  available} The plots of the left shape show that much of the
preffered locations for the proteins to maximize are similar between
this experimental and simulated shape.  The maxima in the top center
of the cell does not seem to appear in the experimental version.  Also
in both of the simulations on the left there don't seem to be maxima
at membrane regions with negative radii of curvature while the
experimental shows a maxima in the bottom center region right at one
of these regions.  The shape on the right compares less successfully.
Here the simulation shows a limited number of maxima locations that
appear to be quite regular, where mannik's experimental results show a
much more chaotic system.  In particular, the maxima location on the
right wall seems to be very robust.  It acts as one of the poles do in
the pill shapes.  \fixme{I'm now running a sim with a slightly
  different shape to see result} The simulated version has found a
robust two pole pattern, similar to that seen in other shapes.
